---
title: "Analytics Workflow for BikeShare"
output:
  html_document: default
  word_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```

# Overview

Summary includes an overiew of the analysis and necessary references.

### For the first leg of value proposition (when to advertise, registered v casual)
* Classification analysis for which days attract most riders
* drill down on whether these are casual or registered riders

### For the second leg (How many bikes to put out each day to ease maintenance)
* multiple regression on number of riders at a given date/time

```{r, echo=FALSE}
## Manage Packages
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(readxl,fpp2,ggplot2,scales,dplyr, skimr, DataExplorer)
```

```{r, echo=FALSE}
#Reading in data
setwd("M:/ISA 616/ISA616BikeShareAnalysis/BikeShare Workflow")
bikeshare <- read.csv("Bike Share Data.csv", stringsAsFactors = TRUE)

#Splitting date and time column into a separate date column and numeric time column
bikeshare$Date <- sapply(strsplit(as.character(bikeshare$datetime), " "), "[", 1)
bikeshare$Time <- sapply(strsplit(as.character(bikeshare$datetime), " "), "[", 2)
bikeshare$Time <- sapply(strsplit(as.character(bikeshare$datetime), " "), "[", 2)
bikeshare$Date <- as.Date(bikeshare$datetime,format='%m/%d/%Y')
bikeshare$Time <- as.factor(bikeshare$Time)
bikeshare$Time <- as.numeric(bikeshare$Time)

#recoding other variables as factors 
bikeshare$season <- factor(bikeshare$season)
bikeshare$holiday <- factor(bikeshare$holiday)
bikeshare$workingday <- factor(bikeshare$workingday)
bikeshare$weather <- factor(bikeshare$weather)

#Removing original datetime column
library(dplyr)
bikeshare<-select(bikeshare, -datetime)
```

# Data Description

The source of this data set is the Capital BikeShare system which tracks bike sharing programs and includes relevant attributes such as seasonal and weather information and membership level. The data analyzed includes the total count of riders by hour and day, along with these additional attributes for bike rentals between the years 2011 and 2012.

* **Number of Observations:** _10,886_

*	**season:**
    + 1 = spring 
    + 2 = summer 
    + 3 = fall
    + 4 = winter. Categorical
*	**holiday:** whether the day is considered a holiday. Categorical. 2 levels
*	**workingday:** whether the day is neither a weekend nor holiday Categorical. 2 levels
*	**weather:** 
    + 1 = Clear, Few clouds, Partly cloudy, Partly cloudy 
    + 2 = Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist 
    + 3 = Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds
    + 4 = Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog 

*	**temp:** temperature in Celsius numerical
*	**atemp:** "feels like" temperature in Celsius numerical
*	**humidity:** relative humidity numerical
*	**windspeed:** wind speed numerical
*	**casual:** number of non-registered user rentals initiated numerical
*	**registered:** number of registered user rentals initiated
*	**count:** number of total rentals (Note that casual+registered=count, so you shouldn’t use casual and registered to predict count. I have included the breakdown in case you wish to predict casual and registered with separate models.)
*	**date:** the current day, formatted as a date.
*	**time:** the current time (in hours) numerical

```{R} 
#DATA DESCRIPTION Code
source('data summary.R')
data.summary(bikeshare)

head(bikeshare)
tail(bikeshare)

```
# Preprocessing


* Details on data preprocessing are provided
* Feauture Generation: talk about creating dummies for multi level factors
* Imputation: No imputation because no missing values
* Outlier Removal: did not remove outliers yet (3% of data) because it is in second half of year, decent amount 
* Cleaning or merging of categories: had to split the date and time column earlier
* Anything that changes original form


```{R, echo=FALSE}
#Creating Dummy variables for multilevel Factors

dum<-as.data.frame(model.matrix(~0+bikeshare$season))
colnames(dum)<-c("spring", "summer", "fall", "winter")
bikeshare<-cbind(bikeshare, dum[,-1])


dum4<-as.data.frame(model.matrix(~0+bikeshare$weather))
colnames(dum4)<-c("Clear", "Mist", "Light Snow/Rain", "Heavy Rain/Snow")
bikeshare<-cbind(bikeshare, dum4[,-1])

bikeshare<-select(bikeshare, -season)
bikeshare<-select(bikeshare, -weather)

```

# Model Selection and Output


```{r}

```
# Justification of anaysis

```{r}

```

# Conclusion

```{r}

```

## References

```{r}


```


