---
title: "Analytics Workflow for BikeShare"
author: "John Marot"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    number_sections: FALSE
    code_folding: hide
  word_document: default

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```

# Overview

Within this document I will detail the analysis that I have completed to assist the theoretical company DownTown Bikes (DTB). I have provided the company with a model to predict the the total count of bike rentals each day. My analysis also offers DTB the opportunity to drill down and predict the number of casual (non-registered) rentals per day, as well as what variables are most important in predicting this value. 

Due to maintenance costs and risk mitigation, it is advantageous for the DTB operations team to better predict the necessary amount of bikes to make available each day. To solve this problem, I utilized multiple regression analysis so that DTB can manage inventory more efficiently. Additionally, DTB's marketing team is attempting to convert more casual riders into registered riders. Using my model, the marketing team can better predict which days attract the most casual riders and tailor promotions based on this.


### Package Management and Loading Data
The below code contains package management using the library pacman. If pacman is required and is not yet installed, it will install the package. Then by using pacman one can call all the required packages and they will be the same version as used in this analysis.   

```{r}
if(require(pacman)==FALSE) install.packages("pacman")
pacman::p_load(readxl,fpp2,ggplot2,scales,dplyr, forecast, rsample, rpart, rpart.plot, ipred, caret, DT )

#Reading in data
setwd("M:/ISA 616/ISA616BikeShareAnalysis/BikeShare Workflow")
bikeshare <- read.csv("Bike Share Data.csv", stringsAsFactors = TRUE)
```

# Data Description

The source of this data set is the Capital BikeShare system which tracks over 500 different bike sharing programs and includes relevant attributes such as seasonal and weather information. The data analyzed includes the total count of riders by hour and day, along with these additional attributes for bike rentals between the years 2011 and 2012.

* **Number of Observations:** _10,886_
* **Missing Observations:** _0_

### Variables

*	**datetime:** hourly date + timestamp. Categorical, but will need to be split into two diferent numerical variables

*	**season:**
    + 1 = spring 
    + 2 = summer 
    + 3 = fall
    + 4 = winter. Categorical. 4 levels
*	**holiday:** whether the day is considered a holiday. Categorical. 2 levels
*	**workingday:** whether the day is neither a weekend nor holiday. Categorical. 2 levels
*	**weather:** 
    + 1 = Clear, Few clouds, Partly cloudy, Partly cloudy 
    + 2 = Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist 
    + 3 = Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds
    + 4 = Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog. Categorical. 4 levels

*	**temp:** temperature in Celsius. Numerical
*	**atemp:** "feels like" temperature in Celsius. Numerical
*	**humidity:** relative humidity. Numerical
*	**windspeed:** wind speed. Numerical
*	**casual:** number of non-registered user rentals initiated. Numerical
*	**registered:** number of registered user rentals initiated. Numerical
*	**count:** number of total rentals initiated (casual + registered).  Numerical


The following code includes a call for a data summary function and prints the head and tail of the data set.

```{R} 
#Calling data summary function that is included within the github.
source('data summary.R')
data.summary(bikeshare)
```

##### Head of the data frame: 

```{R} 
DT::datatable(head(bikeshare, 10))
```

##### Tail of the data frame:

```{R} 
DT::datatable(tail(bikeshare, 10))
```

# Preprocessing

The first step was to separate datetime into a separate date column and numeric time column. This will allow for the inclusion of either date or time, if one wanted to perform a time series analysis on the data. 


```{R} 
bikeshare$Date <- sapply(strsplit(as.character(bikeshare$datetime), " "), "[", 1)
bikeshare$Time <- sapply(strsplit(as.character(bikeshare$datetime), " "), "[", 2)
bikeshare$Date <- as.Date(bikeshare$datetime,format='%m/%d/%Y')
bikeshare$Time <- as.factor(bikeshare$Time)
bikeshare$Time <- as.numeric(bikeshare$Time)

#Removing original datetime using select function in dplyr
bikeshare<-select(bikeshare, -datetime)
```

Next, it was necessary to re-code the variables season, holiday, workingday, and weather as factors. Following this, I added dummy variables for weather and season and appended these to the data frame.

```{R}
#recoding other variables as factors 
bikeshare$season <- factor(bikeshare$season) 
bikeshare$holiday <- factor(bikeshare$holiday)
bikeshare$workingday <- factor(bikeshare$workingday)
bikeshare$weather <- factor(bikeshare$weather)

#creating dummy variables for weather
dum1<-as.data.frame(model.matrix(~0+bikeshare$weather))
colnames(dum1)<-c("Clear", "Mist", "Light_SnowRain", "Heavy_SnowRain")
bikeshare<-cbind(bikeshare, dum1[,-1])

dum<-as.data.frame(model.matrix(~0+bikeshare$season))
colnames(dum)<-c("Spring", "Summer", "Fall", "Winter")
bikeshare<-cbind(bikeshare, dum[,-1])

#Removing original weather variable
bikeshare<-select(bikeshare, -weather)
bikeshare<-select(bikeshare, -season)
```

In terms of outlier removal, 3% of the observations were outliers. The outliers do not appear to be incorrectly measured data and do not affect model assumptions. Because of the relevance and variance provided from these observations I have decided to keep the outliers.  


```{R, echo=FALSE} 
boxplot(bikeshare$count)
```
# Model Selection and Output

To avoid overfitting, I have partitioned the data. The code below involves setting the seed and splitting the data into the training set (70%  of data) and the validation set (30% of data).

```{r}
set.seed(13)
trainIndex = sample(1:nrow(bikeshare), size = round(0.7*nrow(bikeshare)), replace=FALSE)

bikeshare.train<-bikeshare[trainIndex, ]
bikeshare.valid<-bikeshare[-trainIndex, ]

```

## Multiple Regression Analysis


The multiple regression model will be used to predict the overall count of riders, which is the sum of casual and registered rides. Therefore we need to take casual and registered out of the data frame for the analysis because they cannot be predictors for count. 

```{r}
bikeshare.train<-select(bikeshare.train, -registered)
bikeshare.train<-select(bikeshare.train, -casual)
bikeshare.valid<-select(bikeshare.valid, -registered)
bikeshare.valid<-select(bikeshare.valid, -casual)
```

In addition to this, the data includes time variables such as date, time, and quarter. By looking at the plot below, one can see that there is significant monthly seasonality when looking at daily counts. To prevent the seasonality from affecting the model, I have removed the variables related to time.

```{r, echo=FALSE}
bikeshare.train<-select(bikeshare.train, -Time)
bikeshare.train<-select(bikeshare.train, -Date)
bikeshare.train<-select(bikeshare.train, -Summer)
bikeshare.train<-select(bikeshare.train, -Fall)
bikeshare.train<-select(bikeshare.train, -Winter)
bikeshare.valid<-select(bikeshare.valid, -Time)
bikeshare.valid<-select(bikeshare.valid, -Date)
bikeshare.valid<-select(bikeshare.valid, -Summer)
bikeshare.valid<-select(bikeshare.valid, -Fall)
bikeshare.valid<-select(bikeshare.valid, -Winter)
```

The below code includes the initial fitting of the model to all the available variables. The step-wise function then executes subsequently, using this full model as its upper range. Finally, the step-wise model outputs the optimal multiple regression equation and the values of the coefficients. 

```{r}
#First fitting model with all variables available in training set
full<-lm(count~ .,data=bikeshare.train)


#Next using step-wise function to create multiple regression for total Count
null<-lm(bikeshare.train$count~1, data=bikeshare.train)
lm.step<-step(null, scope=list(lower=null, upper=full), direction="both", trace=0)

summary(lm.step)
```
### Multiple Regression Evaluation

Using a stepwise function yielded a model that uses humidity, Mist, atemp, and windspeed to predict the count of rentals. The adjusted R-squared for the model is 0.24, which shows no overfitting, yet is relatively low. Nonetheless, using a stepwise algorithm certifies that the model is using the optimal predictors based on a sequence of f-tests, so attempting to improve the adjusted R-squared is unneccessary. One can see that this model has an RMSE of 158.82 which means that on average, we would expect the error of a prediction using this model to be off by 158.82 daily bike rentals. This may seem like a large error, yet one must consider that some days have thousands of rentals.

```{r, echo=FALSE}
#Model Evaluation
p.full<-predict(full, newdata=bikeshare.valid)
p.step<-predict(lm.step, newdata=bikeshare.valid)

accuracy(p.step, bikeshare.valid$count)
```


## Regression Tree Analysis

The regression tree analysis was done to drill down into the count of casual riders. 


```{r, echo=FALSE}
set.seed(13)
trainIndex1 = sample(1:nrow(bikeshare), size = round(0.7*nrow(bikeshare)), replace=FALSE)

str(bikeshare)

bikeshare.train1<-bikeshare[trainIndex1, ]
bikeshare.valid1<-bikeshare[-trainIndex1, ]

bikeshare.train1<-select(bikeshare.train1, -count)
bikeshare.valid1<-select(bikeshare.valid1, -count)
bikeshare.train1<-select(bikeshare.train1, -registered)
bikeshare.valid1<-select(bikeshare.valid1, -registered)

```

Creating regression tree algorithm.

```{r}

m1 <- rpart(
  formula = casual ~ ., 
  data    = bikeshare.train1,
  method  = "anova"
)
rpart.plot(m1)
```
| 

Cross validating the regression tree with the cv bagged model which allows me to cross validate among different folds of the training data.

```{r}
ctrl <- trainControl(method = "cv",  number = 10) 

# CV bagged model
bagged_cv <- train(
  casual ~ holiday + workingday + temp + humidity + windspeed,
  data = bikeshare.train1,
  method = "treebag",
  trControl = ctrl,
  importance = TRUE
)

# assess results
bagged_cv
```

Plotting the regressions tree determinance of importance among the different variables in the model.


```{r}
# plot most important variables
plot(varImp(bagged_cv)) 
```
### Regression Tree Evaluation

As one can see, humidity, temperature, and workingday are the variables with the larest amount of importance within the regression tree analysis. Using this analysis to supplement the previous multiple regression analysis for the casual count will allow the marketing team at DTB to have a better understanding of which days they should expect a large amount of casual bikers within the city. Additionally, according to the RMSE, one can predict that when using the regression tree the prediction for casual rentals will have an average error of 33.15 rentals.


# Conclusion

In all, DTB has the opportunity to utilize the multiple regression model to improve upon operations and advertising. One considerable takeaway is the relationship between humidity and rentals. When holding the other variables constant, a one unit increase in the humidity will result in an average decrease of 2.789 riders. Within the regression tree analysis, one can see the association between humidity and casual rentals as well. This is shown with humidity being ranked as the most important variable in that analysis. It is interesting to note that a split on humidity of 50 in the regression tree caused a 45% difference in the number of casual rentals. Using this information, DTB may want to caution putting an excess amount of bikes out during extremely humid weather. Also, if DTB is looking to convert casual customers through promotions, they will want to avoid advertising these promotions on non-working days with high humidity and temperatures because the model predicts a low amount of casual rentals. Both strategies will allow for cost savings, as well as gains due to improved bike maintenance, customer retention, and more profitable customers. 


