---
title: "Analytics Workflow for BikeShare"
output:
  html_document: default
  word_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = TRUE, 
                      warning = FALSE,
                      message = FALSE)
```

# Overview

Summary includes an overiew of the analysis and necessary references.

### For the first leg of value proposition (when to advertise, registered v casual)
* Classification analysis for which days attract most riders
* drill down on whether these are casual or registered riders

### For the second leg (How many bikes to put out each day to ease maintenance)
* multiple regression on number of riders at a given date/time

```{r}
setwd("M:/ISA 616/ISA616BikeShareAnalysis/BikeShare Workflow")
bikeshare <- read.csv("Bike Share Data.csv", stringsAsFactors = TRUE)
```

# Data Description
Summary includes a complete description of the dat
```{r}

```

# Preprocessing
Details on data preprocessing are provided
```{R}

```

# Model Selection and Output


```{r}

```
# Justification of anaysis

```{r}

```

# Conclusion

```{r}

```

## References

```{r}


```


